Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2019-04-26 paper/paper_final_version_JVI_no-figs-supp.tex
created version with no figs / supp to convert to Word

\documentclass[]{asm-article}

\usepackage{blindtext}

\usepackage[version=4]{mhchem}
\usepackage{siunitx}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% additional packages imported by Jesse Bloom
\usepackage[export]{adjustbox}
\usepackage{float}

% Added by Jesse Bloom for figures
\renewcommand{\floatpagefraction}{.9}
\setcounter{topnumber}{8}
\setcounter{bottomnumber}{8}
\setcounter{totalnumber}{8}

\newfloat{suppfile}{thp}{lofsupfile}
\floatname{suppfile}{\bf FILE}
\renewcommand{\thesuppfile}{S\arabic{suppfile}}

\newfloat{suppfig}{thp}{lofsupfile}
\floatname{suppfig}{\bf FIG}
\renewcommand{\thesuppfig}{S\arabic{suppfig}}

\renewcommand{\figureautorefname}{Fig.}
\newcommand{\FIG}[1]{\autoref{fig:#1}}

\newcommand{\suppfileautorefname}{File}
\newcommand{\SUPPFILE}[1]{\autoref{suppfile:#1}}

\newcommand{\suppfigautorefname}{Fig.}
\newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}

% Allow big figures on pages
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.9}

\title{Single-cell virus sequencing of influenza infections that trigger innate immunity}

\author{%
Alistair B. Russell,\afn{a}
Elizaveta Elshina,\afn{b}
Jacob R. Kowalsky,\afn{a}
Aartjan J. W. te Velthuis,\afn{b}
Jesse D. Bloom\afn{a,c,d,\authfn{1}}
}

\affil{%
\afn{a}Basic Sciences and Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, United States

\afn{b} Division of Virology, Department of Pathology, University of Cambridge, Cambridge, United Kingdom

\afn{c} Department of Genome Sciences, University of Washington, Seattle, United States

\afn{d} Howard Hughes Medical Institute, Seattle, United States
}

\runningauthor{Russell et al.}
\runningtitle{Single-cell virus sequencing of influenza infections that trigger innate immunity}

\corraddress{Jesse D. Bloom, jbloom@fredhutch.org}

\papertype{Research Article}

@@

\maketitle

\begin{abstract}
Influenza-infected cells vary widely in their expression of viral genes, and only occasionally activate innate immunity.
Here we develop a new method to assess how the genetic variation in viral populations contributes to this heterogeneity.
We do this by determining the transcriptome and full-length sequences of all viral genes in single cells infected with a nominally ``pure'' stock of influenza virus.
Most cells are infected by virions with defects, some of which increase the frequency of innate-immune activation.
These immunostimulatory defects are diverse, and include mutations that perturb the function of the viral polymerase protein PB1, large internal deletions in viral genes, and failure to express the virus's interferon antagonist NS1.
However, immune activation remains stochastic in cells infected by virions with these defects, and occasionally is triggered even by virions that express unmutated copies of all genes.
Our work shows that the diverse spectrum of defects in influenza virus populations contributes to---but does not completely explain---the heterogeneity in viral gene expression and immune activation in single infected cells.

\begin{importance}
Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
But so far, it has been impossible to fully characterize the sequence of the virion infecting any given cell, since conventional techniques like flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
Here we develop a new approach that uses long-read PacBio sequencing to determine the sequences of virions infecting single cells.
We show that viral genetic variation explains some but not all of the cell-to-cell variability in viral gene expression and innate-immune induction.
Overall, our study provides the first complete picture of how viral mutations affect the course of infection in single cells.
\end{importance}

\end{abstract}

\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; defective virus; heterogeneity}

\section{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.

@@

\section{Results}

\subsection{A system to identify and enrich rare IFN+ cells.}

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.4\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.03\textwidth}
{\bf \Large B} \includegraphics[width=0.26\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
Reporter cells to identify and enrich infections that activate IFN expression.
(A) The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
(B)
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

Influenza virus only rarely triggers IFN expression in infected cells~\cite{killip2017single, russell2018extreme, kallfass2013visualizing}---a fact that poses a challenge for the study IFN induction in single cells.
Therefore, we developed a method to identify and enrich rare IFN+ cells by creating A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporters can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by infection with a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN'') directly from reverse-genetics plasmids~\cite{hoffmann2000dna}, and passaged this stock at low multiplicity of infection (MOI).
This process ensures that the viral stock that is relatively ``pure'', with only low levels of the large internal deletions and other defects that arise in stocks passaged at high MOI~\cite{xue2016propagation}.
As described in the next subsection, our stock actually consisted of a mix of two viruses: wild-type WSN and a variant of this virus that carries synonymous viral ``barcodes'' near the termini of each gene.
This viral stock activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
We also validated that MACS for the cell-surface protein driven by the IFN reporter enriched the IFN+ cells by $>$50-fold (\SUPPFIG{MACS}).

\subsection{Combined transcriptomics and virus-sequencing of single infected cells.}

We developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated the viral stock described in the previous subsection, which consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells (this moderately low MOI reasonably balances our desire to limit the number of co-infections with the cost of performing transcriptomics on uninfected cells).
From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
To ensure the presence of IFN-negative cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

%%% start workflow figure
\begin{figure}
\begin{fullwidth}

\includegraphics[width=\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}

\caption{
Approach for combined transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
(A)
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS, and pooled with non-enriched cells and uninfected canine cells that serve as a control for multiplets and mRNA leakage.
(B)
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
(C)
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
(D)
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are sequenced by PacBio).
(E)
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

\end{fullwidth}
\end{figure}
%%% end workflow figure

We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to viral as well as cellular mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs)~\cite{travers2010flexible}.
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells.}

%%% start transcriptomics figure
\begin{figure}
\begin{fullwidth}

\includegraphics[width=\linewidth, clip=false]{figures/single_cell_figures/p_cell_summary.pdf}

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
(A) 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are derived from multiple cells. 
(B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
(C) Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
(D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell derived from each gene.
(F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
(G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
}
\label{fig:transcriptomics}

\end{fullwidth}
\end{figure}
%%% end transcriptomics figure

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\cite{haque2017practical}.
After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B)

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that 162 of 290 infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work using the WSN strain~\cite{russell2018extreme}, and slightly to substantially higher than that reported in studies by others using different viral strains or methodologies~\cite{heldt2015single, brooke2013most, dou2017analysis, jacobs2019incomplete}.

The amount of viral mRNA was lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection complements otherwise transcriptionally defective virions to increase the likelihood that we identify a cell as infected.
The first explanation seems unlikely, as there is no tendency for co-infected cells to express more IFN (\SUPPFIG{coinf_IFN}). 
Therefore we favor the latter two explanations, both of which have been demonstrated for other viruses~\cite{aguilera2017plaques, combe2015single}.
The moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
Over 20\% of infected cells were IFN+, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
The expression of type I and type III IFN was highly correlated in single cells, justifying our decision to collapse both classes under the single label of ``IFN'' in the analyses that follow (\SUPPFIG{type_I_III_correlation}).
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that activated IFN in response to non-viral ligands~\cite{dhir2018mitochondrial, benitez2015vivo, chiang2018viral} or because some cells that we classified as uninfected were actually infected at low levels.
The difference in the frequency of IFN positivity among infected and uninfected cells in \FIG{transcriptomics}G was highly significant ($P < 10^{-5}$, Fisher's exact test).
Many more cells expressed ISGs than IFN itself (\SUPPFIG{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
These results are consistent with the established knowledge that IFN is expressed primarily in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\cite{stetson2006type,honda2006type}.

Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\cite{maaten2008visualizing} of the cells.
Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.

\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells.}

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.85\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
Viral genotypes and infection outcomes in single cells.
Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
For instance, \textit{cell 5} was co-infected by a virion with one unmutated and one internally deleted copy of PB1.
}
\label{fig:genotypes}

\end{fullwidth}
\end{figure}
%%% end genotypes figure %%%

We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes in single infected cells.
We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

>>>>

\documentclass[]{article}

\input{asm-biblatex.tex}
\bibliography{references}

@@

Single-cell virus sequencing of influenza infections that trigger innate immunity

Alistair B. Russell,$^{a}$
Elizaveta Elshina,$^{b}$
Jacob R. Kowalsky,$^{a}$
Aartjan J. W. te Velthuis,$^{b}$
Jesse D. Bloom$^{a,c,d,1}$

$^{a}$ Basic Sciences and Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, United States

$^{b}$ Division of Virology, Department of Pathology, University of Cambridge, Cambridge, United Kingdom

$^{c}$ Department of Genome Sciences, University of Washington, Seattle, United States

$^{d}$ Howard Hughes Medical Institute, Seattle, United States

$^1$ Jesse D. Bloom, jbloom@fredhutch.org

Running title: Single-cell virus sequencing of influenza infections that trigger innate immunity

Keywords: influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; defective virus; heterogeneity

\section*{Abstract}
Influenza-infected cells vary widely in their expression of viral genes, and only occasionally activate innate immunity.
Here we develop a new method to assess how the genetic variation in viral populations contributes to this heterogeneity.
We do this by determining the transcriptome and full-length sequences of all viral genes in single cells infected with a nominally ``pure'' stock of influenza virus.
Most cells are infected by virions with defects, some of which increase the frequency of innate-immune activation.
These immunostimulatory defects are diverse, and include mutations that perturb the function of the viral polymerase protein PB1, large internal deletions in viral genes, and failure to express the virus's interferon antagonist NS1.
However, immune activation remains stochastic in cells infected by virions with these defects, and occasionally is triggered even by virions that express unmutated copies of all genes.
Our work shows that the diverse spectrum of defects in influenza virus populations contributes to---but does not completely explain---the heterogeneity in viral gene expression and immune activation in single infected cells.

\section*{Importance}
Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
But so far, it has been impossible to fully characterize the sequence of the virion infecting any given cell, since conventional techniques like flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
Here we develop a new approach that uses long-read PacBio sequencing to determine the sequences of virions infecting single cells.
We show that viral genetic variation explains some but not all of the cell-to-cell variability in viral gene expression and innate-immune induction.
Overall, our study provides the first complete picture of how viral mutations affect the course of infection in single cells.

\section*{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.

@@

\section*{Results}

\subsection*{A system to identify and enrich rare IFN+ cells.}

Influenza virus only rarely triggers IFN expression in infected cells~\cite{killip2017single, russell2018extreme, kallfass2013visualizing}---a fact that poses a challenge for the study IFN induction in single cells.
Therefore, we developed a method to identify and enrich rare IFN+ cells by creating A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (Fig. 1A).
Cells that activate the IFN reporters can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by infection with a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (Fig. S1A), and activation of the type I and type III IFN reporters was highly correlated in our cells (Fig. S1B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN'') directly from reverse-genetics plasmids~\cite{hoffmann2000dna}, and passaged this stock at low multiplicity of infection (MOI).
This process ensures that the viral stock that is relatively ``pure'', with only low levels of the large internal deletions and other defects that arise in stocks passaged at high MOI~\cite{xue2016propagation}.
As described in the next subsection, our stock actually consisted of a mix of two viruses: wild-type WSN and a variant of this virus that carries synonymous viral ``barcodes'' near the termini of each gene.
This viral stock activated the IFN reporter in $\sim$0.5\% of infected cells  (Fig. 1B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
We also validated that MACS for the cell-surface protein driven by the IFN reporter enriched the IFN+ cells by $>$50-fold (Fig. S3).

\subsection*{Combined transcriptomics and virus-sequencing of single infected cells.}

We developed the approach in Fig. 2 to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated the viral stock described in the previous subsection, which consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (File S2).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (Fig. 2A) at a dose that led to detectable viral transcription in $\sim$25\% of cells (this moderately low MOI reasonably balances our desire to limit the number of co-infections with the cost of performing transcriptomics on uninfected cells).
From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
To ensure the presence of IFN-negative cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (Fig. 2B).
Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to viral as well as cellular mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (Fig. 2C).
But we also took a portion and enriched for full-length viral molecules by PCR (Fig. 2D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs)~\cite{travers2010flexible}.
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (Fig. 2E).

\subsection*{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells.}

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (Fig. 3A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\cite{haque2017practical}.
After this filtering, we retained 1,490 human cells for further analysis (Fig. 3B)

To identify infected cells, we examined the fraction of each transcriptome derived from virus (Fig. 3C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (Fig. 3C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in Fig. 3C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
Fig. 3D (top panel) shows that 162 of 290 infected cells express all eight genes (see Fig. S4 for frequencies for individual genes).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work using the WSN strain~\cite{russell2018extreme}, and slightly to substantially higher than that reported in studies by others using different viral strains or methodologies~\cite{heldt2015single, brooke2013most, dou2017analysis, jacobs2019incomplete}.

The amount of viral mRNA was lower in cells that failed to express viral genes (Fig. 3D, bottom panel).
However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (Fig. 3D, bottom panel).
Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (Fig. 3E) and generally matched values from older bulk studies~\cite{hatada1989control}.

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (Fig. 3F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From Fig. 3F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (Fig. 3C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection complements otherwise transcriptionally defective virions to increase the likelihood that we identify a cell as infected.
The first explanation seems unlikely, as there is no tendency for co-infected cells to express more IFN (Fig. S5). 
Therefore we favor the latter two explanations, both of which have been demonstrated for other viruses~\cite{aguilera2017plaques, combe2015single}.
The moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.

We next examined expression of IFN and ISGs (Fig. 3G and Fig. S6).
Over 20\% of infected cells were IFN+, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
The expression of type I and type III IFN was highly correlated in single cells, justifying our decision to collapse both classes under the single label of ``IFN'' in the analyses that follow (Fig. S7).
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that activated IFN in response to non-viral ligands~\cite{dhir2018mitochondrial, benitez2015vivo, chiang2018viral} or because some cells that we classified as uninfected were actually infected at low levels.
The difference in the frequency of IFN positivity among infected and uninfected cells in Fig. 3G was highly significant ($P < 10^{-5}$, Fisher's exact test).
Many more cells expressed ISGs than IFN itself (Fig. S6A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (Fig. S6B).
These results are consistent with the established knowledge that IFN is expressed primarily in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\cite{stetson2006type,honda2006type}.

Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
Fig. S8 shows unsupervised t-SNE clustering~\cite{maaten2008visualizing} of the cells.
Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.

\subsection*{Full genotypes of viruses infecting single IFN+ and IFN- cells.}

We next used PacBio sequencing (Fig. 2D) to determine the full sequences of the viral genes in single infected cells.
We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (Fig. S9).
The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (Fig. S10), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
